---
title: 'Inference for numerical data'
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---


```{r global-options, include=FALSE}
library(dplyr)
library(ggplot2)
library(oilabs)
library(haven)
data(nc)
dat = nc
dat = as.data.frame(dat)
colnames(dat) = gsub("[.]", "_", colnames(dat))
write_dta(data = dat, path = "nc.dta", version = 13)

library(statamd)
statapath = statamd::stata_engine_path()
profile_do(dataset = "nc.dta")

knitr::opts_chunk$set(eval = FALSE, engine = "stata", 
    engine.path = statapath, comment = "")
```

## Getting Started

### The data

In 2004, the state of North Carolina released a large data set containing 
information on births recorded in this state. This data set is useful to 
researchers studying the relation between habits and practices of expectant 
mothers and the birth of their children. We will work with a random sample of 
observations from this data set.

Load the `nc` data set into our workspace.

```{r load-data}
use "nc.dta"
```

We have observations on 13 different variables, some categorical and some 
numerical. Each observation is a birth recorded in North Carolina.


1.  How many cases are there in our sample?

Remember that you can answer this question by viewing the data in the Data Editor or by looking at the bottom right in the *Properties* window.



## Exploratory data analysis

We will first start with analyzing the weight gained by mothers throughout the 
pregnancy: `gained`.

Using visualization and summary statistics, describe the distribution of weight 
gained by mothers during pregnancy. The `summarize` function can be useful.

```{r summary}
summarize gained
```

1.  How many mothers are we missing weight gain data from?  **Hint:** The `codebook` command is useful for finding missing values.

Next, consider the possible relationship between a mother's smoking habit and the 
weight of her baby. Plotting the data is a useful first step because it helps 
us quickly visualize trends, identify strong associations, and develop research
questions.

2.  Make a side-by-side boxplot of `habit` and `weight`. What does the plot 
highlight about the relationship between these two variables?

The box plots show how the medians of the two distributions compare, but we can
also compare the means of the distributions using the following to 
first sort the data by the `habit` variable, and then calculate the mean
`weight` separately for these groups using the `summarize` function.

```{r by-means}
bysort weight: summarize weight
```

There is an observed difference, but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis test.

## Inference

3.  Are all conditions necessary for inference satisfied? Comment on each. You can 
compute the group sizes with the `summarize` command above.

4.  Write the hypotheses for testing if the average weights of babies born to 
smoking and non-smoking mothers are different.


```{r inf-weight-habit-ht, tidy=FALSE}
ttest weight, by(habit)
```

5.  Construct and record a confidence 
interval for the difference between the weights of babies born to nonsmoking and 
smoking mothers, and interpret this interval in context of the data. Note that by 
default you'll get a 95% confidence interval. If you want to change the
confidence level, add the option (`level(X)`) where X is a value
between 0 and 100. 

```{r inf-weight-habit-ht, tidy=FALSE}
ttest weight, by(habit)
```

By default the function reports an interval for the difference `diff` specified as ($\mu_{nonsmoker} - \mu_{smoker}$)
. 

**FIX THIS**

We can change this order by changing the reference category for habit.  using the `order` argument:

```{r inf-weight-habit-ci, tidy=FALSE}
inference(y = weight, x = habit, data = nc, statistic = "mean", type = "ci", 
          method = "theoretical", order = c("smoker","nonsmoker"))
```

* * *

## More Practice

6.  Calculate a 95% confidence interval for the average length of pregnancies 
(`weeks`) and interpret it in context. **Hint:** Try using the function `mean`.

7.  Calculate a new confidence interval for the same parameter at the 90% 
confidence level. You can change the confidence level using the same `level` option discussed above. Comment on the width of this interval versus 
the one obtained in the previous exercise.

8.  Conduct a hypothesis test evaluating whether the average weight gained by 
younger mothers is different than the average weight gained by mature mothers.

9.  Now, a non-inference task: Determine the age cutoff for younger and mature 
mothers. Use a method of your choice, and explain how your method works.

10. Pick a pair of variables: one numerical (response) and one categorical (explanatory). 
Come up with a research question evaluating the relationship between these variables. 
Formulate the question in a way that it can be answered using a hypothesis test
and/or a confidence interval. Answer your question using Stata, report the statistical results, and also provide an explanation in 
plain language. Be sure to check all assumptions, state your $\alpha$ level, and conclude 
in context.

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons 
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0).
This lab was adapted for Stata by Jenna R Krall and OpenIntro by Mine &Ccedil;etinkaya-Rundel from a lab written by the faculty and TAs of UCLA Statistics.
</div>