---
title: "Multiple linear regression"
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
---


```{r global-options, include=FALSE}
library(dplyr)
library(ggplot2)
library(oilabs)
library(haven)
data(evals)
dat = evals
dat = as.data.frame(dat)
colnames(dat) = gsub("[.]", "_", colnames(dat))
write_dta(data = dat, path = "evals.dta", version = 13)

library(statamd)
statapath = statamd::stata_engine_path()
profile_do(dataset = "evals.dta")

knitr::opts_chunk$set(eval = FALSE, engine = "stata", 
    engine.path = statapath, comment = "")
```


## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" by Hamermesh and Parker 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. 

In this lab we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.

## Getting Started

### The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. The result is a data frame 
where each row contains a different course and columns represent variables about 
the courses and professors.

Let's load the data:

```{r load-data, message=FALSE}
use "evals.dta"
```

We have observations on 21 different variables, some categorical and some 
numerical. 

<!--The meaning of each variable can be found by bringing up the help file:
-->



## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrased? If not, rephrase the question.

2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not?

3.  Excluding `score`, select two other variables and describe their relationship 
    with each other using an appropriate visualization.

## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case.  We need the variables `score` and average beauty rating of the professor, `bty_avg`.

```{r scatter-score-bty_avg}
twoway scatter score bty_avg
```

Before we draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

4.  Replot the scatterplot, but this time use the option `jitter(5)`. What was
    misleading about the initial scatterplot?

```{r scatter-score-bty_avg-jitter}
twoway scatter score bty_avg, jitter(5)
```

5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model to predict average
    professor score (`score`) by average beauty rating. Write out the equation for the linear 
    model and interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?
    
Add the line of the best fit model to your plot using the following:
    
```{r scatter-score-bty_avg-line-se}
twoway scatter score bty_avg || lfit score bty_avg
```

The red line is the model, indicated in the legend as "Fitted values". 


6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel}
twoway scatter bty_avg bty_f1lower
cor bty_avg bty_f1lower
```

As expected the relationship is quite strong---after all, the average score is 
calculated using the individual scores. We can actually look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels}
graph matrix bty_f1lower bty_f1upper bty_f2upper bty_m1lower bty_m1upper bty_m2upper bty_avg
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after we've accounted for the gender of the professor, we can add the gender 
term into the model.

```{r scatter-score-bty_avg_gender}
regress score bty_avg i.gender
```

Note that we use `i.gender` because gender is a categorical variable.

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.

8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?

Note that the row in the regression output for `gender` is now called `gender male`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that Stata 
recodes `gender` from having the values of `female` and `male` to being an 
indicator variable called `gender male` that takes a value of $0$ for females and
a value of $1$ for males (Such variables are often referred to as "dummy" 
variables.).

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]



9.  What is the equation of the line corresponding to males? (*Hint:* For
    males, the parameter estimate is multiplied by 1.) For two professors
    who received the same beauty rating, which gender tends to have the
    higher course evaluation score?

The decision to call the indicator variable `gender male` instead of`gender female`
has no deeper meaning. Stata simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using `ibX.gender` in place of `i.gender` in the model, where `X` is the desired reference level. In the case of gender, there are two categories: (1) male and (2) female.  Therefore, to change the reference group to females,

```{r recodegender}
regress score bty_avg ib2.gender
```


10. Create a new model with `gender` removed and `rank` 
    added in. How does Stata appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.

The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors teaching the course, number of credits, average beauty rating, outfit, and 
picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.

Let's run the model:

```{r m_full, tidy = FALSE}
regress score rank ethnicity gender language age cls_perc_eval cls_students  cls_level cls_profs cls_credits bty_avg pic_outfit pic_color
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.

13. Interpret the coefficient associated with the ethnicity variable.

14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.  

16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.

17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?

18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.

19. Would you be comfortable generalizing your conclusions to apply to professors
    generally (at any university)? Why or why not?

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). This lab was adapted for Stata by Jenna R Krall from a lab written by 
Mine &Ccedil;etinkaya-Rundel and Andrew Bray.
</div>

