---
title: "Probability"
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---

```{r global-options, include=FALSE}
library(dplyr)
library(ggplot2)
library(oilabs)
library(haven)
data(kobe_basket)
dat = kobe_basket
dat = as.data.frame(dat)
colnames(dat) = gsub("[.]", "_", colnames(dat))
write_dta(data = dat, path = "kobe_basket.dta", version = 13)

library(statamd)
statapath = statamd::stata_engine_path()
profile_do(dataset = "kobe_basket.dta")


# kobe streak
kobe_streak <- calc_streak(kobe_basket$shot)
dat = kobe_streak
dat = as.data.frame(dat)
colnames(dat) = gsub("[.]", "_", colnames(dat))
write_dta(data = dat, path = "kobe_streak.dta", version = 13)

# add in idependent streaks
set.seed(4400)
shot_outcomes <- c("H", "M")
sim_basket <- sample(shot_outcomes, size = 133, replace = TRUE, prob = c(0.45, 0.55))
ind_streak <- calc_streak(sim_basket)
dat = ind_streak
dat = as.data.frame(dat)
colnames(dat) = gsub("[.]", "_", colnames(dat))
write_dta(data = dat, path = "ind_streak.dta", version = 13)


knitr::opts_chunk$set(eval = FALSE, engine = "stata", 
                      engine.path = statapath, comment = "")
```

## The Hot Hand

Basketball players who make several baskets in succession are described as 
having a *hot hand*. Fans and players have long believed in the hot hand 
phenomenon, which refutes the assumption that each shot is independent of the 
next. However, [a 1985 paper](http://www.sciencedirect.com/science/article/pii/0010028585900106) by Gilovich, Vallone, and Tversky collected evidence
that contradicted this belief and showed that successive shots are independent 
events. This paper started a great controversy that continues to this day, as you can 
see by Googling *hot hand basketball*.

We do not expect to resolve this controversy today. However, in this lab we'll 
apply one approach to answering questions like this. The goals for this lab are 
to (1) think about the effects of independent and dependent events, (2) learn 
how to simulate shooting streaks in R, and (3) to compare a simulation to actual
data in order to determine if the hot hand phenomenon appears to be real.

## Getting Started


### Data

Our investigation will focus on the performance of one player: [Kobe Bryant](https://en.wikipedia.org/wiki/Kobe_Bryant) of 
the Los Angeles Lakers. His performance against the Orlando Magic in the [2009 
NBA Finals](https://en.wikipedia.org/wiki/2009_NBA_Finals) earned him the title *Most Valuable Player* and many spectators 
commented on how he appeared to show a hot hand. Let's load some necessary files
that we will need for this lab.  Specifically, read in the dataset `kobe_basket.dta` into Stata.

This data frame contains 133 observations and 6 variables, where every
row records a shot taken by Kobe Bryant. The `shot` variable in this dataset
indicates whether the shot was a hit (`H`) or a miss (`M`).

Just looking at the string of hits and misses, it can be difficult to gauge 
whether or not it seems like Kobe was shooting with a hot hand. One way we can 
approach this is by considering the belief that hot hand shooters tend to go on 
shooting streaks. For this lab, we define the length of a shooting streak to be 
the *number of consecutive baskets made until a miss occurs*.

For example, in Game 1 Kobe had the following sequence of hits and misses from 
his nine shot attempts in the first quarter:

\[ \textrm{H M | M | H H M | M | M | M} \]

You can verify this by viewing the first 8 rows of the data in the *Data Browser*.

Within the nine shot attempts, there are six streaks, which are separated by a 
"|" above. Their lengths are one, zero, two, zero, zero, zero (in order of 
occurrence).

1.  What does a streak length of 1 mean, i.e. how many hits and misses are in a 
    streak of 1? What about a streak length of 0?

Counting streak lengths manually for all 133 shots would get tedious, so we'll
use another dataset, `kobe_streak.dta` to view all the streaks from the `kobe_basket` dataset.

```{r calc-streak-kobe}
use "kobe_streak.dta"
```

We can then take a look at the distribution of these streak lengths.

```{r plot-streak-kobe, eval = FALSE, echo = 2}
use "kobe_streak.dta", clear
qnorm length
```


2.  Describe the distribution of Kobe's streak lengths from the 2009 NBA finals. 
    What was his typical streak length? How long was his longest streak of 
    baskets? Make sure to include the accompanying plot in your answer.

## Compared to What?

We've shown that Kobe had some long shooting streaks, but are they long enough 
to support the belief that he had a hot hand? What can we compare them to?

To answer these questions, let's return to the idea of *independence*. Two 
processes are independent if the outcome of one process doesn't effect the outcome 
of the second. If each shot that a player takes is an independent process, 
having made or missed your first shot will not affect the probability that you
will make or miss your second shot.

A shooter with a hot hand will have shots that are *not* independent of one 
another. Specifically, if the shooter makes his first shot, the hot hand model 
says he will have a *higher* probability of making his second shot.

Let's suppose for a moment that the hot hand model is valid for Kobe. During his
career, the percentage of time Kobe makes a basket (i.e. his shooting 
percentage) is about 45%, or in probability notation,

\[ P(\textrm{shot 1 = H}) = 0.45 \]

If he makes the first shot and has a hot hand (*not* independent shots), then 
the probability that he makes his second shot would go up to, let's say, 60%,

\[ P(\textrm{shot 2 = H} \, | \, \textrm{shot 1 = H}) = 0.60 \]

As a result of these increased probabilites, you'd expect Kobe to have longer 
streaks. Compare this to the skeptical perspective where Kobe does *not* have a
hot hand, where each shot is independent of the next. If he hit his first shot,
the probability that he makes the second is still 0.45.

\[ P(\textrm{shot 2 = H} \, | \, \textrm{shot 1 = H}) = 0.45 \]

In other words, making the first shot did nothing to effect the probability that
he'd make his second shot. If Kobe's shots are independent, then he'd have the 
same probability of hitting every shot regardless of his past shots: 45%.

Now that we've phrased the situation in terms of independent shots, let's return
to the question: how do we tell if Kobe's shooting streaks are long enough to 
indicate that he has a hot hand? We can compare his streak lengths to someone
without a hot hand: an independent shooter. 

## Simulations in Stata

While we don't have any data from a shooter we know to have independent shots, 
that sort of data is very easy to simulate in Stata. In a simulation, you set the 
ground rules of a random process and then the computer uses random numbers to 
generate an outcome that adheres to those rules.

As a simple example, you can
simulate flipping a fair coin.  We first will clear out the existing data from Stata.  We can always read the data back in the future.

```{r}
clear
```

Then, we set how many samples we want to take, for example 100.  

```{r}
set obs 100
```

If want to sample "heads" or "tails", we could set "heads" as 1 and "tails" as 0.  This is sampling from a binomial distribution, so we use the `rbinomial` function.  The arugments of the function correspond to having each observation represent a coin flip (1), and that the coin is fair (0.5).

```{r head-tail}
generate sample = rbinomial(1, 0.5)
```

The function `rbinomial` can be thought of as a hat with two slips of paper in it: 
one slip says `heads` and the other says `tails`. The function draws 
one slip from the hat and tells us if it was a head (1) or a tail (0). 

Look at the variable `sample` in the *Data Browser*.  Just like when flipping a 
coin, sometimes you'll get a heads, sometimes you'll get a tails, but in the 
long run, you'd expect to get roughly equal numbers of each.




To view the results of this simulation, use
`table` to count up the number of heads and tails.

```{r table-sim-fair-coin, echo = 4}
clear
set obs 100
generate sample = rbinomial(1, 0.5)
table sample
```

Say we're trying to simulate an unfair coin 
that we know only lands heads 20% of the time. We can adjust for this by changing the second argument in `rbinomial`.

```{r sim-unfair-coin}
generate sample2 = rbinomial(100, 0.2)
```

This indicates that for the two options (heads and tails),
we want to select the first one, `heads`, with probability 0.2 and the second 
one, `tails` with probability 0.8. Another way of thinking about this is to 
think of the outcome space as a bag of 10 chips, where 2 chips are labeled 
"head" and 8 chips "tail". Therefore at each draw, the probability of drawing a 
chip that says "head"" is 20%, and "tail" is 80%.

3.  In your simulation of flipping the unfair coin 100 times, how many flips 
    came up heads? Include the code for sampling the unfair coin in your response.
    Since your Stata do-file will generate a new sample each time
    you run it, you should also "set a seed" **before** you sample. Read more 
    about setting a seed below.
    
<div id="boxedtext">
**A note on setting a seed:** Setting a seed will cause Stata to select the same 
sample each time you run your code. This will make sure your results don't
change each time you run your code, and it will also ensure reproducibility of your work 
(by setting the same seed it will be possible to reproduce your results). You can
set a seed like this:
```{r set-seed}
set seed 35797                  # make sure to change the seed
```
The number above is completely arbitraty. If you need inspiration, you can use your
ID, birthday, or just a random string of numbers. The important thing is that you
use each seed only once. Remember to do this **before** you sample in the exercise 
above.
</div>

In a sense, we've shrunken the size of the slip of paper that says "heads", 
making it less likely to be drawn and we've increased the size of the slip of 
paper saying "tails", making it more likely to be drawn. When we simulated the 
fair coin, both slips of paper were the same size. 

If you want to learn more about `rbinomial` or any other function, recall that you 
can always check out its help file.

```{r help-sample,tidy = FALSE}
help rbinomial
```

## Simulating the Independent Shooter

Simulating a basketball player who has independent shots uses the same mechanism 
that we use to simulate a coin flip. It is still choosing between two options: ("Miss", "Hit").

To make a valid comparison between Kobe and our simulated independent shooter, 
we need to align both their shooting percentage and the number of attempted shots.  We first need to read in our dataset `kobe_basket.dta` again.  Notice we can apply the option `clear` to clear our simulated data.

```{r, kobe-baskets2}
use "kobe_basket.dta", clear
```

Now we need to generate a new variable that is our simulated shots, assuming no hot hand.

4.  What change needs to be made to the `rbinomial` function so that it reflects a 
    shooting percentage of 45%? Make this adjustment, then run a simulation to 
    sample 133 shots. Note that because we already have a dataset loaded, we do not need to set the observations using `set obs`.  Assign the output of this simulation to a new variable called
    `sim_basket`.

With the results of the simulation saved as `sim_basket`, we have the data 
necessary to compare Kobe to our independent shooter.

Both variables, `shot` and `sim_basket` represent the results of 133 shot attempts, each with the same 
shooting percentage of 45%. We know that our simulated data is from a shooter 
that has independent shots. That is, we know the simulated shooter does not have
a hot hand.

* * *

## More Practice

### Comparing Kobe Bryant to the Independent Shooter

5.  Using the dataset `ind_streak.dta`, describe the distribution of streak lengths.  Make sure
    to include a plot in your answer.

6.  What is the typical streak 
    length for this simulated independent shooter with a 45% shooting percentage?
    How long is the player's longest streak of baskets in 133 shots? 

7.  If you were to run the simulation of the independent shooter a second time, 
    how would you expect its streak distribution to compare to the distribution 
    from the question above? Exactly the same? Somewhat similar? Totally 
    different? Explain your reasoning.

8.  How does Kobe Bryant's distribution of streak lengths compare to the 
    distribution of streak lengths for the simulated shooter? Using this 
    comparison, do you have evidence that the hot hand model fits Kobe's 
    shooting patterns? Explain.


</div>

<div id="license">
This is a product of OpenIntro that is released under a 
[Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). 
This lab was adapted for Stata by Jenna R. Krall and John Muschelli and adapted for OpenIntro by Andrew Bray and Mine &Ccedil;etinkaya-Rundel
from a lab written by Mark Hansen of UCLA Statistics.
</div>