---
title: 'Foundations for statistical inference - Confidence intervals'
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---



```{r global-options, include=FALSE}
library(dplyr)
library(ggplot2)
library(oilabs)
library(haven)
data(ames)
dat = ames
dat = as.data.frame(dat)
colnames(dat) = gsub("[.]", "_", colnames(dat))
write_dta(data = dat, path = "ames.dta", version = 13)

library(statamd)
statapath = statamd::stata_engine_path()
profile_do(dataset = "ames.dta")

knitr::opts_chunk$set(eval = FALSE, engine = "stata", 
    engine.path = statapath, comment = "")
```


If you have access to data on an entire population, say the size of every 
house in Ames, Iowa, it's straightforward to answer questions like, "How big 
is the typical house in Ames?" and "How much variation is there in sizes of 
houses?". If you have access to only a sample of the population, as is often 
the case, the task becomes more complicated. What is your best guess for the 
typical size if you only know the sizes of several dozen houses? This sort of 
situation requires that you use your sample to make inference on what your 
population looks like.

<div id="boxedtext">
**Setting a seed:** We will take some random samples and build sampling distributions
in this lab, which means you should set a seed on top of your lab. If this concept
is new to you, review the lab concerning probability.
</div>

## Getting Started


### The data

We consider real estate data from the city of Ames, Iowa. This is the same 
dataset used in the previous lab. The details of 
every real estate transaction in Ames is recorded by the City Assessor's 
office. Our particular focus for this lab will be all residential home sales 
in Ames between 2006 and 2010.  This collection represents our population of 
interest. In this lab we would like to learn about these home sales by taking 
smaller samples from the full population. Let's load the data.

```{r load-data}
use "ames.dta"
```

In this lab we'll start with a simple random sample of size 60 from the 
population. Do not forget to `preserve` the original dataset within your Stata session.  See lab 5 for more information.

```{r sample}
preserve
sample 60, count
```

Note that 
the data set has information on many housing variables, but for the first 
portion of the lab we'll focus on the size of the house, represented by the 
variable `area`.

1.  Describe the distribution of house area in your sample. What would you say is the
    "typical" size within your sample? Also state precisely what you interpreted 
    "typical" to mean.

1.  Would you expect another student's distribution to be identical to yours? 
    Would you expect it to be similar? Why or why not?

## Confidence intervals

Return for a moment to the question that first motivated this lab: based on 
this sample, what can we infer about the population? Based only on this single 
sample, the best estimate of the average living area of houses sold in Ames 
would be the sample mean, usually denoted as $\bar{x}$ (here we're calling it 
`x_bar`). That serves as a good **point estimate** but it would be useful 
to also communicate how uncertain we are of that estimate. This uncertainty
can be quantified using a **confidence interval**.

A confidence interval for a population mean is of the following form
\[ \bar{x} + z^\star \frac{s}{\sqrt{n}} \]

You should by now be comfortable with calculating the mean and standard deviation of 
a sample in Stata. And we know that the sample size is 60. So the only remaining building
block is finding the appropriate critical value for a given confidence level. We can
use the `invnormal` function for this task, which will give the critical value associated
with a given percentile under the normal distribution. Remember that confidence levels
and percentiles are not equivalent. For example, a 95% confidence level refers to the
middle 95% of the distribution, and the critical value associated with this area will
correspond to the 97.5th percentile.

We can find the critical value for a 95% confidence interal using
```{r z_star_95}
display invnormal(0.975)
```
which is roughly equal to the value critical value 1.96 that you're likely
familiar with by now.  Recall that `display` makes Stata perform as a calculator.

Let's finally calculate the confidence interval using the equation above.  Use `summarize` to find the sample size, sample mean, and sample standard deviation.


To recap: even though we don't know what the full population looks like, we're 95% 
confident that the true average size of houses in Ames lies between the lower and upper endpoints of your confidence interval. There are a few conditions that must be met for this interval to be valid.

1.  For the confidence interval to be valid, the sample mean must be normally 
    distributed and have standard error $s / \sqrt{n}$. What conditions must be 
    met for this to be true?

## Confidence levels

1.  What does "95% confidence" mean?

In this case we have the rare luxury of knowing the true population mean since we 
have data on the entire population. Let's calculate this value so that
we can determine if our confidence intervals actually capture it. 

```{r pop-mean}
restore, preserve
summarize area
```

1.  Does your confidence interval capture the true average size of houses in 
    Ames? If you are working on this lab in a classroom, does your neighbor's 
    interval capture this value? 

1.  Each student should have gotten a slightly different confidence interval. What 
    proportion of those intervals would you expect to capture the true population 
    mean? Why?

Using Stata, we're going to collect many samples to learn more about how sample 
means and confidence intervals vary from one sample to another.

First, let's introduce the function `mean`, which gives the sample mean, standard error, and 95\% confidence interval for the population mean.

```{r mean-fun}
mean area
```

Here is the rough outline:

-   Obtain a random sample of 60.
-   Calculate the sample's mean and standard deviation
-   Repeat these steps 50 times and save the dataset as `confint.dta`
-   Use these means and standard deviations to calculate the lower and upper bounds of the confidence intervals.

We can accomplish this using the `bootstrap` function that we introduced in lab 5. The following lines of 
code takes 50 random samples of size `n` from population (and remember we defined 
$n = 60$ earlier), and computes the upper and lower bounds of the confidence intervals based on these samples.

```{r calculate-50-cis}
bootstrap mean=r(mean) sd=r(sd), saving("confint.dta", replace) size(60) reps(50): summarize area
use "confint.dta", clear
generate se = sd / sqrt(60)
generate lower = mean - 1.96 * se
generate upper = mean + 1.96 * se
```

Open up the Data Editor to view the first five intervals:

Next we'll create a plot similar to Figure 4.8 on page 175 of [OpenIntro Statistics, 3rd
Edition](https://www.openintro.org/os). The first step will be to create a new variable in 
the `ci` data frame that indicates whether the interval does or does not capture the 
true population mean. Note that capturing this value would mean the lower bound of the
confidence interval is below the value and upper bound of the confidence interval is
above the value. 

```{r capture-mu}
generate capturemu = "no" 
replace capturemu = "yes" if lower < 1499.69 & upper > 1499.69
```

You'll notice we created the new variable, `capturemu` in two steps.  First, we set `capturemu` to "no" for all observations.  Then, we set `capturemu` = "yes" for observations where the lower bound was less than the population mean and the upper bound was greater than the population mean, or observations where the interval contains the population mean.

We now have all the information we need to create the plot. 
 

```{r plot-ci}
qplot(data = ci, x = replicate, y = x_bar, color = capture_mu) +
  geom_errorbar(aes(ymin = lower, ymax = upper)) + 
  geom_hline(data = params, aes(yintercept = mu), color = "darkgray") + # draw vertical line
  coord_flip()
```

1.  What proportion of your confidence intervals include the true population mean? Is 
    this proportion exactly equal to the confidence level? If not, explain why. Make 
    sure to include your plot in your answer.

* * *

## More Practice

1.  Pick a confidence level of your choosing, provided it is not 95%. What is 
    the appropriate critical value?

1.  Calculate 50 confidence intervals at the confidence level you chose in the 
    previous question, and plot all intervals on one plot, and calculate the proportion 
    of intervals that include the true population mean. How does this percentage compare 
    to the confidence level selected for the intervals? Make 
    sure to include your plot in your answer.

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons 
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0).
This lab was adapted for Stata by Jenna R Krall and written for OpenIntro by Andrew Bray and Mine &Ccedil;etinkaya-Rundel.
</div>